---
execute:
  warning: false
format:
  html:
    embed-resources: true
    code-fold: true
---

# Data cleaning

```{r setup-cleaning}
# packages
library(tidyverse)
library(readxl)
library(ggmap)
library(janitor)
library(ggrepel)
library(GGally)
library(targets)
library(gt)
library(DataExplorer)
library(patchwork)
library(ggh4x)
library(ggthemes)
library(glmmTMB)
library(DHARMa)
library(performance)
library(MASS)
library(scales)
library(viridis)
library(gtsummary)
library(ggeffects)
library(multcomp)
library(broom)
select <- dplyr::select

# set theme
survey_theme <- theme_few()
theme_set(survey_theme)

## for renaming sample ids to fix inconsistencies
mareano_renamer <- function(ss_in) {
  
  sub_renamer <- function(ss_in){
  # check input
  stopifnot("requires single character string input." = length(ss_in) == 1 && is.character(ss_in))

  # for all strings remove spaces and replace BC with BX
  ss_out <- ss_in %>%
    str_replace_all("\\s", "") %>%
    str_replace_all("BC", "BX")
  # return if string do not contain MC or BX, as errors only with BX/MC
  if (str_detect(ss_out, "MC|BX", negate = TRUE)) {
    return(ss_out)}
  
  # part of name including BX/MC
  ss_prefix <- str_extract(ss_out, ".+[BXMC]+")
  # suffix-numbers with incoherent lengths
  ss_suffix <- str_extract(ss_out, "[0-9]+$")

  # change lengths
  if (str_length(ss_suffix) == 3) {
    return(ss_out)
  } else if (str_length(ss_suffix) == 2) {
    return(paste0(ss_prefix, "0", ss_suffix))
  } else if (str_length(ss_suffix) == 1) {
    return(paste0(ss_prefix, "00", ss_suffix))
  } else {
    # break with error message "suffix string not of length 1, 2 or 3":
    stop("suffix string not of length 1, 2 or 3")
  }
  }
  
  ss_out <-  map_chr(ss_in, sub_renamer)
  return(ss_out)
}


# to fix shifted column names
newnames <- (read_excel("data/Hele datasettet v2.xlsx", skip = 178) %>% clean_names())[1, 16:29] %>% as_vector()

# appending correct column names
dataset <- read_excel("data/Hele datasettet v2.xlsx", skip = 178) %>%
  clean_names() %>%
  rename_with(.cols = x16:x29, ~newnames) %>%
  clean_names() %>%
  slice(-1) %>% 
  select(
    sample_id = sample,
    year = cruise_3,
    number = cruise_4,
    from_cm_o = sample_interval_top_bottom_5,
    to_cm_o = x6,
    longitude = dde,
    latitude = ddn,
    depth_m = mbsl,
    sample_id2 = x1096mc002,
    sample_id3 = sample_2,
    from_cm_i = sample_interval_top_bottom_12,
    to_cm_i = x13,
    clay_fraction,
    silt_fraction,
    clay_silt_fraction = clay_silt_percent,
    toc = toc_percent,
    pfos:pf_tri_a,
    -pf_dc_s,
    sum_9_pfas = sum_10_pfas,
  ) %>%
  filter(
    !is.na(year)
  ) %>%
  mutate(across(!matches("sample"), as.numeric)) %>% 
    #assign areas:
  mutate(area = case_when(
    row_number() %in% c(1:18) ~ "SV",
    row_number() %in% c(19:45) ~ "BH mid",
    row_number() %in% c(46:66) ~ "BH sør",
    row_number() %in% c(67:89) ~ "NH",
    row_number() %in% c(90:95) ~ "S",
    TRUE ~ "Other"
  ),
  area = as.factor(area)) %>% 
  pivot_longer(
    cols = pfos:sum_9_pfas, names_to = "pfas", values_to = "pfas_value"
  ) %>%
  mutate(sample_id3 = mareano_renamer(sample_id3))


# import curated LOQs
loqs_raw <- read_excel("data/LOQs.xlsx") %>%
  clean_names() %>%
  select(-enhet) %>%
  rename(pfas = forbindelse)

loqs_years <- loqs_raw %>%
  filter(!(is.na(pfas))) %>%
  mutate(across(-pfas, ~ str_replace_all(.x, "\\<", "") %>% as.numeric()),
    pfas = make_clean_names(pfas)
  ) %>%
  rename_with(.cols = everything(), ~ str_replace(.x, "_2", "A") %>% str_replace("_3", "B")) %>%
  pivot_longer(cols = matches("\\d{4}"), names_to = "analysis_year", values_to = "loq")

loqs_samples <- loqs_raw %>%
  filter(is.na(pfas)) %>%
  rename_with(.cols = everything(), ~ str_replace(.x, "_2", "A") %>% str_replace("_3", "B")) %>%
  select(-pfas) %>%
  pivot_longer(
    cols = matches("\\d{4}"),
    names_to = "analysis_year",
    values_to = "sample_id"
  ) %>%
  drop_na() %>% 
  mutate(
    sample_id = mareano_renamer(sample_id)
  )

loqs <- left_join(
  loqs_years, loqs_samples,
  by = c("analysis_year")
)

# merge data, assign values below LOQ
pfas_raw <- left_join(dataset, loqs, by = c("pfas", "sample_id3" = "sample_id")) %>%
  select(-analysis_year) %>%
  select(sample_id3, year, area, longitude, latitude, clay_silt_fraction, toc, pfas, pfas_value, loq)
  
pfas_data_eq0 <- pfas_raw %>%
  group_by(pfas) %>%
  mutate(max_loq = max(loq)) %>%
  mutate(
    pfas_value = case_when(
      pfas_value < max_loq ~ 0,
      TRUE ~ pfas_value
    )
  ) %>%
  ungroup() %>%
  group_by(sample_id3) %>%
  mutate(
    pfas_value = if_else(
      pfas == "sum_9_pfas",
      sum(pfas_value[pfas != "sum_9_pfas"], na.rm = TRUE),
      pfas_value
    )
  ) %>%
  select(-max_loq)

pfas_data_eq0_wide <- pivot_wider(pfas_data_eq0 %>% select(-c(loq)), names_from = pfas, values_from = pfas_value) %>% ungroup()

```




# Todo
-   Add missing longitudes
-   Explain why silt/clay better than TOC for paper.
-   Add text with interpretation
-   Change clay+silt to mud.

# Modeling
<details>

<summary>Model details</summary>

## Sum PFAS: Tweedie LOQ-0 (candidate)


### Fit
```{r tweedie-fit}

fit_tweedie_0 <- glmmTMB((sum_9_pfas) ~ area + clay_silt_fraction, data = pfas_data_eq0_wide, family = tweedie(link = "log"))

fit_tweedie_0 %>% stepAIC()

broom.mixed::glance(fit_tweedie_0) %>% 
  bind_cols(tibble(
  tweedie_power = glmmTMB:::.tweedie_power(fit_tweedie_0),
  tweedie_dispersion = sigma(fit_tweedie_0)
)) %>% gt()

broom.mixed::tidy(fit_tweedie_0) %>% select(-c(1,2)) %>% gt()

```

### Validation

```{r sum-dispersion-test-tweedie}
sim_residuals <- simulateResiduals(fittedModel = fit_tweedie_0, n = 1000)
testDispersion(sim_residuals)
```

```{r sum-tweedie-val}
#| layout-ncol: 2
#| layout-nrow: 5

plotResiduals(sim_residuals, quantreg = TRUE, smoothScatter = FALSE)

plotResiduals(sim_residuals, form = pfas_data_eq0_wide$toc, xlab = "TOC")


plotQQunif(sim_residuals, testUniformity = TRUE, 
           testOutliers = TRUE, testDispersion = TRUE)

plotResiduals(sim_residuals, form = pfas_data_eq0_wide$year, xlab = "Year",)

plotResiduals(sim_residuals, form = pfas_data_eq0_wide$latitude, xlab = "Latitude")

plotResiduals(sim_residuals, form = pfas_data_eq0_wide$longitude, xlab = "Longitude")

plotResiduals(sim_residuals, form = pfas_data_eq0_wide$area, xlab = "Area")
```
Adding a quadratic term decreases the residuals on toc, though increases AIC. Thus, TOC is kept outside model. Spatial dependency appear to not be systematic, no temporal dependency observed.


```{r sum-tweedie-sim}
#| label: fig-tweedie-obs-sim
#| fig-cap: "Black dots shows observed values, fitted a Tweedie GLM with log link, values below LOQ set to 0. Ribbons shows the 95% CI of the mean predicted value. Red dots shows simulated values from 100 simulations based on observed values for the covariates."


# predict values for CIs
pred <- ggpredict(fit_tweedie_0, terms = c("clay_silt_fraction [1:100]", "area"))
# simulate
simulated <- simulate(fit_tweedie_0, nsim = 10, seed = 12345)
sim_data <- bind_cols(pfas_data_eq0_wide, simulated) %>%
  pivot_longer(cols = matches("sim"), names_to = "simgroup", values_to = "sim") %>%
  rename(group = area)

pred_plot <- pred %>% ggplot(aes(x, predicted)) +
  geom_smooth(color = muted("red")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_point(data = sim_data, aes(x = clay_silt_fraction, y = sim), alpha = 0.1, size = 0.2, color = muted("red"))+
  geom_point(aes(x = clay_silt_fraction, y = sum_9_pfas), data = pfas_data_eq0_wide %>% rename(group = area)) +
  facet_wrap(~group)+
  xlab("Clay + silt fraction [%]")+
  ylab("\u03A39PFAS")

pred_plot

```

Model refitted without SV and S retains the clay + silt vs $\Sigma$PFAS relationship.


## 3PFAS: Tweedie LOQ-0 (candidate)


### Fit
```{r 3pf-tweedie-fit}


fit_tweedie_pfos <- glmmTMB((pfos) ~ area + clay_silt_fraction, data = pfas_data_eq0_wide, family = tweedie(link = "log"))

fit_tweedie_pfos %>% summary()

fit_tweedie_pfos %>% stepAIC(direction = "both", scope = list(upper = ., lower = ~ 1))


fit_tweedie_pfoa <- glmmTMB((pfoa) ~ area + clay_silt_fraction, data = pfas_data_eq0_wide, family = tweedie(link = "log"))

fit_tweedie_pfoa %>% summary()

fit_tweedie_pfoa %>% stepAIC(direction = "both", scope = list(upper = ., lower = ~ 1))


fit_tweedie_pfna <- glmmTMB((pfna) ~ area + clay_silt_fraction, data = pfas_data_eq0_wide, family = tweedie(link = "log"))

fit_tweedie_pfna %>% stepAIC(direction = "both", scope = list(upper = ., lower = ~ 1))



map_df(
  list(fit_tweedie_pfos, fit_tweedie_pfoa, fit_tweedie_pfna),
  ~ broom.mixed::glance(.x) %>%
  bind_cols(tibble(
    tweedie_power = glmmTMB:::.tweedie_power(.x),
    tweedie_dispersion = sigma(.x)) %>% 
      mutate(model = str_flatten(.x$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~"), .before = 1)
  )) %>% 
  select(model, AIC, tweedie_power, tweedie_dispersion) %>%
  gt()


map_df(
  list(fit_tweedie_pfos, fit_tweedie_pfoa, fit_tweedie_pfna),
  ~ broom.mixed::tidy(.x, exponentiate = TRUE) %>%
    select(-c(1, 2)) %>%
    mutate(model = str_flatten(.x$frame %>% colnames()) %>% str_extract("pfos|pfoa|pfna"), .before = 1)
) %>% 
  select(model, everything()) %>% 
  gt(groupname_col = "model") %>%
  fmt_number(columns = vars(estimate, std.error, statistic, p.value), decimals = 3) %>% 
  # if p.value below 0.5, make text font bold:
  tab_style(style = list( 
                         cell_text(weight = 'bold')), 
            locations = cells_body(columns=vars(p.value), 
                                   rows = p.value <= 0.05)
  ) %>% 
  #make groupname_col bold:
  tab_style(style = list( 
                         cell_text(weight = 'bold')), 
            locations = cells_column_labels(columns=everything())
  ) %>% 
  tab_style(
  style = cell_text(weight = "bold"),
  locations = cells_row_groups(groups = everything())
) %>% 
  cols_label(estimate = "exp(estimate)")


```


PFOS - clay/silt + area, validation OK (outliers)

PFOA: clay/silt + toc^2^!

PFNA: clay/silt + toc^2^!




### Validation


```{r tweedie-val}
#| layout-ncol: 2
#| layout-nrow: 5

lapply(list(fit_tweedie_pfos, fit_tweedie_pfoa, fit_tweedie_pfna), function(.) {
  
paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")) %>% print()
  
  sim_residuals <- simulateResiduals(fittedModel = ., n = 1000, plot = FALSE)
  
  plotQQunif(.)
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")))
  
  plotResiduals(.)
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")," (Tweedie power=", glmmTMB:::.tweedie_power(.) %>% round(digits = 3), ", sigma=", sigma(.) %>% round(digits = 3),")"))
  
    plotResiduals(., form = pfas_data_eq0_wide$toc, .lab = "TOC")
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")))

    plotResiduals(., form = pfas_data_eq0_wide$year, .lab = "Year")
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")))
  
      plotResiduals(., form = pfas_data_eq0_wide$latitude, .lab = "Latitude")
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")))
  
        plotResiduals(., form = pfas_data_eq0_wide$longitude, .lab = "Longitude")
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")))
  
          plotResiduals(., form = pfas_data_eq0_wide$area, .lab = "Area")
  mtext(paste0(str_flatten(.$frame %>% colnames(), collapse = "+") %>% str_replace("\\+", "~")), side = 1)
  

})


# predict values for CIs
pred <- ggpredict(fit_tweedie_pfos, terms = c("clay_silt_fraction [1:100]", "area"))
# simulate
simulated <- simulate(fit_tweedie_pfos, nsim = 10, seed = 12345)
sim_data <- bind_cols(pfas_data_eq0_wide, simulated) %>%
  pivot_longer(cols = matches("sim"), names_to = "simgroup", values_to = "sim") %>%
  rename(group = area)

pred_plot_pfos <- pred %>% ggplot(aes(x, predicted)) +
  geom_smooth(color = muted("red")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_point(data = sim_data, aes(x = clay_silt_fraction, y = sim), alpha = 0.1, size = 0.2, color = muted("red"))+
  geom_point(aes(x = clay_silt_fraction, y = pfos), data = pfas_data_eq0_wide %>% rename(group = area)) +
  facet_wrap(~group)+
  xlab("Clay + silt fraction [%]")+
  ylab("PFOS")

pred_plot_pfos

pred <- ggpredict(fit_tweedie_pfoa, terms = c("clay_silt_fraction [1:100]", "area"))
# simulate
simulated <- simulate(fit_tweedie_pfoa, nsim = 10, seed = 12345)
sim_data <- bind_cols(pfas_data_eq0_wide, simulated) %>%
  pivot_longer(cols = matches("sim"), names_to = "simgroup", values_to = "sim") %>%
  rename(group = area)

pred_plot_pfoa <- pred %>% ggplot(aes(x, predicted)) +
  geom_smooth(color = muted("red")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_point(data = sim_data, aes(x = clay_silt_fraction, y = sim), alpha = 0.1, size = 0.2, color = muted("red"))+
  geom_point(aes(x = clay_silt_fraction, y = pfoa), data = pfas_data_eq0_wide %>% rename(group = area)) +
  facet_wrap(~group)+
  xlab("Clay + silt fraction [%]")+
  ylab("PFOA")

pred_plot_pfoa

pred <- ggpredict(fit_tweedie_pfna, terms = c("clay_silt_fraction [1:100]", "area"))
# simulate
simulated <- simulate(fit_tweedie_pfna, nsim = 10, seed = 12345)
sim_data <- bind_cols(pfas_data_eq0_wide, simulated) %>%
  pivot_longer(cols = matches("sim"), names_to = "simgroup", values_to = "sim") %>%
  rename(group = area)

pred_plot_pfna <- pred %>% ggplot(aes(x, predicted)) +
  geom_smooth(color = muted("red")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_point(data = sim_data, aes(x = clay_silt_fraction, y = sim), alpha = 0.1, size = 0.2, color = muted("red"))+
  geom_point(aes(x = clay_silt_fraction, y = pfna), data = pfas_data_eq0_wide %>% rename(group = area)) +
  facet_wrap(~group)+
  xlab("Clay + silt fraction [%]")+
  ylab("PFNA")

pred_plot_pfna
```



```{r}
#| label: fig-sum-tweedie-obs-sim
#| fig-cap: "Black dots shows observed values, fitted a Tweedie GLM with log link, values below LOQ set to 0. Ribbons shows the 95% CI of the mean predicted value. Red dots shows simulated values from 100 simulations based on observed values for the covariates."


# predict values for CIs
pred <- ggpredict(fit_tweedie_0, terms = c("clay_silt_fraction [1:100]", "area"))
# simulate
simulated <- simulate(fit_tweedie_0, nsim = 10, seed = 12345)
sim_data <- bind_cols(pfas_data_eq0_wide, simulated) %>%
  pivot_longer(cols = matches("sim"), names_to = "simgroup", values_to = "sim") %>%
  rename(group = area)

pred_plot <- pred %>% ggplot(aes(x, predicted)) +
  geom_smooth(color = muted("red")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_point(data = sim_data, aes(x = clay_silt_fraction, y = sim), alpha = 0.1, size = 0.2, color = muted("red"))+
  geom_point(aes(x = clay_silt_fraction, y = sum_9_pfas), data = pfas_data_eq0_wide %>% rename(group = area)) +
  facet_wrap(~group)+
  xlab("Clay + silt fraction [%]")+
  ylab("\u03A39PFAS")
pred_plot

```

Model refitted without SV and S retains the clay + silt vs $\Sigma$PFAS relationship.

</details>






</details>



# Visualization


```{r map}
#| column: screen
#| out-width: 100%

map_area_conc <-
  ggmap(tar_read(stadia_map)) +
  geom_point(aes(x = longitude, y = latitude, color = sum_9_pfas, fill = sum_9_pfas, shape = area), data = pfas_data_eq0_wide, position = position_dodge(width = 10), size = 2, alpha = 0.7) +
  labs(x = "Longitude [°E]", y = "Latitude [°N]") +
  scale_color_viridis(option = "magma") +
  scale_fill_viridis(option = "magma") +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  #change legend label to Area and \u03A39PFAS:
  labs(color = "\u03A39PFAS", shape = "Area")+
  guides(fill = FALSE)
  

map_area_conc

```


```{r}
ggpairs(pfas_data_eq0_wide %>% 
          select(
            area, clay_silt_fraction, toc, sum_9_pfas, pfos, pfoa, pfna
          ))

ggplot(pfas_data_eq0_wide, aes(clay_silt_fraction, toc))+geom_point()+geom_smooth(method = "lm")+
  xlab("Clay + silt fraction [%]")+
  ylab("TOC")
```



```{r}
pred_plot
```



```{r}
pfas_prop <- pfas_data_eq0 %>%
  # for each area, calculate the proportion of each pfas:
  group_by(area, pfas) %>% 
  summarise(
    pfas_sum = sum(pfas_value, na.rm = TRUE),
  ) %>% 
  group_by(area) %>% 
  mutate(
    pfas_prop = pfas_sum / sum(pfas_sum[pfas != "sum_9_pfas"], na.rm = TRUE)
  ) %>% 
  filter(pfas != "sum_9_pfas")

ggplot(pfas_prop, aes(fill=pfas, y=pfas_prop, x=area)) + 
    geom_bar(position="stack", stat="identity")+
  scale_fill_brewer(palette="Set3")+
  xlab("Area")+ylab("Propotion")+
  #legend title PFAS:
  labs(fill = "PFAS")
```


# Tables

```{r area-hypothesis}

glht(fit_tweedie_0, linfct = mcp(area = "Tukey")) %>%
  # automatically Tukey-adjusted p-values in summary
  summary() %>%  tidy() %>% select(-c(term, null.value)) %>% 
  mutate(estimate = exp(estimate)) %>% 
  rename("exp(estimate)" = estimate) %>% 
  gt() %>% tab_header(title = "Sum9: Tukey post-hoc test for area") %>% 
  fmt_number(columns = 2:5,decimals = 2)


glht(fit_tweedie_pfos, linfct = mcp(area = "Tukey")) %>%
  # automatically Tukey-adjusted p-values in summary
  summary() %>%  tidy() %>% select(-c(term, null.value)) %>% 
  mutate(estimate = exp(estimate)) %>% 
  rename("exp(estimate)" = estimate) %>% 
  gt() %>% tab_header(title = "PFOS: Tukey post-hoc test for area") %>% 
  fmt_number(columns = 2:5,decimals = 2)

glht(fit_tweedie_pfoa, linfct = mcp(area = "Tukey")) %>%
  # automatically Tukey-adjusted p-values in summary
  summary() %>%  tidy() %>% select(-c(term, null.value)) %>% 
  mutate(estimate = exp(estimate)) %>% 
  rename("exp(estimate)" = estimate) %>% 
  gt() %>% tab_header(title = "PFOA: Tukey post-hoc test for area") %>% 
  fmt_number(columns = 2:5,decimals = 2)

glht(fit_tweedie_pfna, linfct = mcp(area = "Tukey")) %>%
  # automatically Tukey-adjusted p-values in summary
  summary() %>%  tidy() %>% select(-c(term, null.value)) %>% 
  mutate(estimate = exp(estimate)) %>% 
  rename("exp(estimate)" = estimate) %>% 
  gt() %>% tab_header(title = "PFNA: Tukey post-hoc test for area") %>% 
  fmt_number(columns = 2:5,decimals = 2)

```


```{r n-below-lod}
# pfas_raw %>% group_by(pfas) %>% 
#   summarise(
#     n = sum(above_loq),
#     n_samples = n_distinct(sample_id3)
#   ) %>% 
#   mutate(
#     p = n / n_samples
#   ) %>% arrange(p) %>% 
#   gt() %>% 
#   fmt_percent(columns = vars(p)) %>% 
#   # name gt table:
#   tab_header(
#     title = md("Number of samples above LOQ"),
#     subtitle = md("Number of samples above LOQ for each PFAS, and the proportion of samples above LOQ.")
#   ) %>% 
#   cols_label(
#     pfas = md("**PFAS**"),
#     n = md("**# samples > LOQ**"),
#     n_samples = md("**# samples**"),
#     p = md("**Proportion > LOQ**")
#   )
```




```{r}

pfas_data_eq0_wide %>% group_by(area) %>% 
  #mean, sd for each area for each of sum_9_pfas, toc, clay_silt_fraction:
  summarise(
    sum_9_pfas = mean(sum_9_pfas, na.rm = TRUE),
    toc = mean(toc, na.rm = TRUE),
    clay_silt_fraction = mean(clay_silt_fraction, na.rm = TRUE),
    n = n()
  ) %>% gt() %>% 
  fmt_number(columns = 2:4,decimals = 1) %>% 
  tab_spanner(label = "Mean", columns = 2:4) %>% 
  tab_header(title = "Summary of covariates by area, <LOQ = 0") %>% 
  cols_label(
    area = md("**Area**"),
    sum_9_pfas = md("**\u03A39PFAS**"),
    toc = md("**TOC**"),
    clay_silt_fraction = md("**Clay + silt %**")
  )

```
